---
title: "Assignment_1"
author: "Shuo Mao 437681258"
date: "2024-07-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Task 1:
## Re-create the following sequences using R without the use of c() and explicit values. Hint: making use of the help files will sometimes help a lot.

(a): 

`(i) [1] -4 4 -4 4 -3 3 -3 2 -2 1`

```{r}
{positive_numbers = rep(4:1, 4:1);#creating numbers
alternating_sequence = positive_numbers * rep(c(-1, 1),#using circle to re-assign their postie or negative sign to it
                                              length.out = length(positive_numbers))}
print(alternating_sequence)
```


`(ii) [1] 1 22 333 4444 55555 666666 7777777 88888888`


```{r}
# using function to create repeated number with no space, then adding space between each repeated numbers
formatted_sequence <- paste(sapply(1:8, function(x) paste(rep(x, x), collapse = ""))# will generate number such as: 1, 22, 333 ....
                            , collapse = " ") # combine those number and re-format it. 

# Print the formatted sequence
cat(formatted_sequence)
```


`(iii) [1] "z" "Y" "x" "W" "v" "U" "t" "S" "r" "Q" "p" "O" "n" "M" "l" "K" "j" "I" "h" "G" [21] "f" "E" "d" "C" "b" "A"`


```{r}

lowercase <- rev(letters)  # "z" to "a"
uppercase <- rev(LETTERS)  # "Z" to "A"

#combine uppercase and lowercase alternatively 
combined <- c(rbind(lowercase[seq(1, 26, 2)], uppercase[seq(2, 26, 2)]))

# Print the combined sequence
print(combined)
```


`(iv) [1] "1^1=1" "2^2=4" "3^3=27" "4^4=256" "5^5=3125" [6] "6^6=46656" "7^7=823543" "8^8=16777216"`



```{r}
# Calculate the power results
{exponents <- 1:8;results <- paste(exponents, "^", exponents, "=", exponents^exponents)}
print(results)
```


`(v) [1] "jan1" "feb2" "mar3" "apr4" "may5" "jun6" "jul7" "aug8" "sep9" "oct10" [11] "nov11" "dec12"`


```{r}
# Create the vector of month abbreviations and corresponding numbers, then Combine the month abbreviations with the numbers
{months <- tolower(month.abb);numbers <- 1:12;result <- paste(months, numbers, sep = "")}
# Print the result
print(result)
```



(b): find and write down six different ways to assign a variable x the value 8

1: using '=' 
```{r}
x = 8 
print(x)
```
2: using '->'
```{r}
8 -> x
print(x)
```

3: using '<-'
```{r}
x <- 8
print(x)
```

4: using assign function
```{r}
assign("x", 8)
print(x)
```

5: using list function 

```{r}
x <<- 8
print(x)
```

6: using c function 

```{r}
x <- c(9,8,1,4)[2]
print(x)

eval(x <- 8)
x
```
## (c) We know that ex = ∑∞ i=0 xi/i!. Approximate this by A(k; x) = ∑k i=0 xi/i! for finite k. Find the value k such that A(k; 1) does not differ by exp(1) on your machine. 


```{r}
# Compute Euler's number
e <- exp(1)

# Initialize variables
k <- 0
approximation <- 0
tolerance <- .Machine$double.eps *2 # Machine precision (default tolerance)

# Loop to find the smallest k
repeat {
  # Update the approximation of A(k; 1)
  approximation <- approximation + 1 / factorial(k)
  
  # Check if the approximation is within the tolerance
  if (abs(approximation - e) <= tolerance) {
    break
  }
  
  # Increment k
  k <- k + 1
}

# Output the result
cat(sprintf("The smallest k such that A(k; 1) approximates exp(1) within machine precision is: %d\n", k))

```

# Task 2:
Consider the data frame xs.nz in VGAMdata. Based on this cross-sectional
study, use R expressions (not your eyes or my eyes) to find the values below (and show the
values only, nothing else). Don’t use any explicit loop or anything equivalent. Ignore (delete)
any cases with NAs in the relevant variables. Your code should be elegant and readable, etc


### (a):The proportion of females who have had an acne problem
```{r}
```
### (b):What is the age of the oldest Maori who has a cat and a dog?
```{r}
```
### (c):mong current smokers, what proportion drinks alcohol each week, on average? What about among nonsmokers?
```{r}
```